Ignition of Fuel Beds by Firebrands

### 2022-05-30 litreview.tex
Updates and added model github link

\chapter{Background}
\label{chap:literature}
    This chapter provides a discussion of the background and motivation of this work in further depth than the manuscripts presented later in this dissertation. Here, a general overview of ignitions as a threat to structures and fire spread, as well as the current state of knowledge and predictive capabilities, is communicated. Each manuscript includes a relevant and more detailed literature review pertaining to the specific objective of the chapter.

@@

\section{The Problem of Ignition}
    Ignition of a home in a WUI fire is attributed to three primary mechanisms; direct flame contact, ignition by radiative heating, and ignition by firebrands. Of these methods, ignition by firebrands is considered to be a primary threat to homes~\cite{Suzuki2021, Mell2010, Manzello2018FORUMResearch}. Case studies of WUI fires have consistently attributed structural losses to firebrand ignition~\cite{Westhaver2017WhyDisaster, Roberts2021} with one study attributing at least 2/3 of home losses to firebrands~\cite{Mell2011}. With the majority of ignitions occurring due to firebrand ignition, it is imperative to understand the mechanism of firebrand ignition to reduce home losses. An analysis of structures burned in California wildfires from 2013 to 2018 found that many of the homes lost would have been considered "fire-safe" by current standards but were still ignited by embers~\cite{Syphard2019}. Codes, standards, and best practices built upon a better understanding of ignition by firebrands would help provide better resilience to firebrand attacks~\cite{Manzello2020} reducing the number of homes lost in wildfires. Examples of these changes may include adjustments in building and landscaping materials favoring the selection of harder to ignite species or housing geometry constraints to reduce the collection of flammable materials or embers. Without a comprehensive knowledge of the ignition process, creating the codes and standards to reduce the risk of home loss is challenging at best and may even be counterproductive.

    The general mechanism for firebrand ignition is considered to be a three-step process that starts within the fire itself, often at a significant distance from the ignition point. First, a firebrand is generated as burning material breaks off or is lofted into the air. Second, the firebrand is then lofted through the air and transported to the ignition site. Third, a firebrand lands on a recipient fuel bed, and then ignition occurs if the criteria for ignition are met~\cite{Babrauskas2003}. When considering the possibility of ignition during a fire event, all three processes must be considered as each process significantly influences the properties of an ember landing on a fuel bed. For example, the number, energy content, and size of embers that may land on a fuel bed are significantly influenced by the surrounding foliage~\cite{Hudson2020EffectsScale} and moisture content of the foliage~\cite{Adusumilli2021FirebrandShrub} which introduces significant variability in ignition sources a fuel bed may encounter. Similarly, the winds transporting a firebrand significantly influence the path of travel and time of flight, which influences the energy content~\cite{Sardoy2007, MatvienkoOVA2016} and localized winds around structures may lead to firebrand accumulation~\cite{Suzuki2020a, Suzuki2017a, Suzuki2015}. Unfortunately, the generation and ignition processes are significantly less understood than the transport processes~\cite{Manzello2020}. The work contained in this dissertation is focused on the ignition process of the fuel bed by a firebrand and does not consider the generation of firebrands. For further information on firebrand generation processes, the reader is referred to review papers considering this process by Fernandez-Pello~\cite{Fernandez-Pello2017}, Manzello et al.~\cite{Manzello2020} and Babrauskas~\cite{Babrauskas2003}.

    Firebrands have been identified as a source of structure loss as early as the 1600s~\cite{Suzuki2020} however, studies of firebrand ignition have only begun recently. Perhaps one of the first studies published to specifically address firebrands within the scope of structure loss was conducted by Waterman in 1969~\cite{Waterman1969ExperiemntalGeneration}. However, research remained sparse until the early 2000s. Early research and a significant portion of current research has been conducted by dropping combusting firebrands~\cite{Ellis2011, Ellis2015, Ganteaume2009, Manzello2006, Manzello2006a} (either flaming or glowing) or hot metal particles~\cite{Wang2017, Urban2017, Fernandez-Pello2015, Hadden2011} onto fuel beds in varying configurations that simulate conditions in a wildfire and observing ignition or no ignition outcomes. These works have identified some parameters that appear to control ignition but models created from these studies only provide qualitative predictions of ignition. The lack of models of quantitative accuracy was suggested by Finney et al. in 2013, who identified the lack of a fundamental understanding of what processes occur and how they interact as a primary gap in the knowledge of ignition~\cite{Finney2013}. A similar need was identified by Manzello et al. in 2018~\cite{Manzello2018} and again in 2020~\cite{Manzello2020}. As fires increase in severity and the WUI continues to expand, the number of homes lost to wildfires is likely to significantly increase in the near future. Thus, the need for accurate predictions of ignition has never been higher.

\section{Modeling Efforts}
    Creating a model with quantitative accuracy of ignition requires a thorough understanding of the processes and relative importance of each process leading up to ignition. Figure~\ref{fig:ignitionDiagram} illustrates the different factors that can influence the ignition process. To better understand how these parameters affect the ignition process can be considered as a series of three processes. The first is the heat transfer to the fuel bed from the ember. The second is the pyrolysis of the fuel bed material and the subsequent release of pyrolyzates into the air above the fuel bed. The third step is the mixing of the pyrolyzates with air above the fuel bed and subsequent gas-phase ignition. It is important to note that all three of these processes may occur simultaneously throughout the ignition process and are often interconnected. For example, as heat is transferred from the firebrand to the fuel bed, the thermal conductivity of the fuel bed changes with temperature both before and after pyrolysis begins~\cite{Fjellerup2003}. Furthermore, the heat transfer mode and thermal resistance between the firebrand and fuel continuously change as mass is released due to pyrolysis, which may influence ignition~\cite{Yang2016EffectParticle}. With the interconnection of the processes that occur during ignition and the inherent variability of fuel beds and biomass fuels in general, the quantification of parameters that influence ignition is quite challenging. Despite these challenges, results of previous research have identified influential parameters that have consistently been observed to control ignition. This work considers these parameters as belonging to one of three categories, namely, thermo-physical properties of the fuel bed, energy supplied by the firebrand, and environmental conditions near the ignition point. Detailed reviews of each category are provided in the relevant manuscripts and are not reproduced here. The remaining sections of this review address the predictive models of ignition that have been created from the current state of knowledge. The accuracy and applicability of each model are discussed in turn, and the remaining challenges are identified as they have motivated the objectives of this work. 
        \begin{figure}[hpbt]
            \centering
                \includegraphics[width=\textwidth]{Figures/dissertationBed.pdf}
            \caption{Illustration of key features of fuel bed ignition and parameters/characteristics that may influence ignition. The black region near the firebrand represents char formed as the result of pyrolysis.}
            \label{fig:ignitionDiagram}
        \end{figure}

@@

        \begin{equation}
            r_{cr} = \delta_{cr}\sqrt{\frac{k}{\rho A \Delta H}\frac{RT^{2}_{p0}}{E} exp\left(\frac{E}{RT_{p0}}  \right)}
            \label{eqn:hotspot}
        \end{equation}
    In this equation, $\delta_{cr}$ is defined as the critical Frank-Kamenetskii hot spot parameter and is determined from calculations~\cite{Goldshleger1973IgnitionDimensions}. $\delta_{cr}$ is dependent on the ratio of volumetric heat capacities of the ignition source and the material, the activation energy of the material, and the temperature of the hot particle. In Equation~\ref{eqn:hotspot} $k, \rho A, E$, and $\Delta H$ are the thermal conductivity, density, pre-exponential factor, activation energy, and heat of combustion of the fuel bed material. $R$ is the universal gas constant, and $T_{p0}$ is the initial temperature of the heat source. Ultimately, Equation~\ref{eqn:hotspot} defines the particle energy needed to raise the temperature of a fuel bed region such that thermal runaway and subsequent ignition occurs. While relatively simple, the hot spot theory has some limitations that prevent quantitative predictions of ignition in a wildfire setting. A discussion of the shortcomings of the hot spot theory is presented by Thomas~\cite{Thomas1965} and more recently presented by Hadden et al.~\cite{Hadden2011} with respect to wildfires. The two most significant shortcomings addressed by the previous works are neglecting heat losses to the surroundings and considering the pyrolysis and ignition phenomena as single-step, infinitely fast reactions. Hadden et al. attribute these shortcomings as a significant source of error for quantitative ignition predictions. This assertion is reinforced in a review by Fernandez-Pello~\cite{Fernandez-Pello2017} who concluded that in instances where hot spot theory has been applied, only qualitatively accurate predictions of ignition had been obtained. A similar model to hot spot theory but with more comprehensive inclusion of heat loss from the ember was developed by Yin et al.~\cite{Yin2014}. A correlation between fuel properties, the energy transferred to the fuel bed, and the time to ignition was proposed and is shown in Equation~\ref{eqn:yinMC}.
        \begin{equation}
            \sqrt{t_{ig}} \approx \frac{q\sqrt{\frac{\rho_{f}k}{c_{p}}}}{\frac{\rho Z \Delta H_c}{t_b}- h_T(T_f - T_0)}
            \label{eqn:yinMC}
        \end{equation}
    Comparing the correlations presented in Equations~\ref{eqn:hotspot} and \ref{eqn:yinMC}, Equation~\ref{eqn:yinMC} includes properties of the fuel bed ($\rho$, $k$, $c_p$, $Z$, $T_0$) and properties the ember or energy source ($\rho_f$, $\Delta H_c/t_b$, $T_f$) as well as the energy required for ignition ($q$) mirroring the Frank-Kamenetskii critical hot spot diameter. The addition of heat losses to the ambient is the most significant difference from the hotspot parameter and addresses a key shortcoming of the hot spot theory. However, the model does not consider the combustion of the fuel bed other than the critical energy for ignition, leaving the second key gap for the hot spot theory unaddressed. Perhaps due to the omission of combustion, this model suffers a similar fate as the hot spot theory and follows qualitative trends but is not quantitative. Ultimately, the hot spot theory and similar models do not capture the complexity and variability inherent to ignition~\cite{Manzello2020}.

@@

    Proposed ignition models that provide greater accuracy than the hot spot and critical mass flux models do so at the expense of complexity and computational cost. The majority of these models and modeling efforts use a two-part approach to join a pyrolysis-specific solver to address the chemical kinetics of an existing or custom computation fluid dynamics solver. This approach requires selecting choosing a combination of three parts, each of which has limitations and considerations which may affect the accuracy of ignition predictions. First, an appropriate chemical kinetic mechanism must be chosen based on the computational resources available. For example, many early works~\cite{DiBlasi1996} and some more recently~\cite{Ding2015, Torero2020} have considered pyrolysis to occur as a single step process or as a process of few reactions (approximately ten or fewer). Newer kinetic mechanisms have been developed to more completely capture the processes of pyrolysis and are much more complex. Perhaps the most comprehensive mechanisms are extended forms of the mechanism developed by Ranzi et al.~\cite{Ranzi2008}. Additions to this mechanism have expanded the method to include extractives to cover a greater variety of biomass~\cite{Debiagi2015} and the addition of reactions to address gas-phase reactions previously not considered~\cite{Dhahak2019}. For a more detailed discussion of the progression and challenges of biomass pyrolysis, the reader is referred to review articles by Di Blasi~\cite{DIBLASI199371} and more recently by Haberle et al.~\cite{Haberle2017}.

    One of the most recent and most promising tools for ignition modeling and pyrolysis is Gpyro which is a pyrolysis solver specifically designed to handle both the condensed (solid) and gas phase of biomass combustion~\cite{Lautenberger2009a}. Gpyro is often coupled with the Fire Dynamics Simulator (FDS)~\cite{FDSManual}, a LES solver specifically designed for physics relevant to fires, to form a complete computational package. The combination of Gpyro and FDS has been utilized to observe combustion sensitivities and ignition sensitivities. For example, a study of the effects of moisture content on ignition using Gpyro/FDS suggested that moisture in the fuel affected ignition, gas-phase combustion, and pyrolysis of the fuel~\cite{Yashwanth2016}. Other studies using Gpyro/FDS have been conducted to predict ignition and observe controlling parameters~\cite{Shotorban2018, Fernandez-Pello2015, Lautenberger2009}. Unfortunately, the Gpyro/FDS modeling requires significant computational expenditure and is not practical to be used for the prediction of ignition across the vast number of conditions and configurations of possible ignition threats in wildland urban interface regions around the globe. Thus, a usable and accurate model will likely land somewhere between the critical mass flux and the full CFD model with respect to complexity. Currently, the understanding of which effects are the most influential to ignition is not complete enough to develop such a model. Bridging this knowledge gap by identifying and quantifying which parameters are the most influential to ignition or non-ignition outcome is the overall goal of this work.

    %     \section{Energy Effects}
    %     Fuel bed response to heating and the impacts that heat transfer has on ignition probability is important knowledge for understanding ignition of fuel beds. Unfortunately, the current state of knowledge can only answer this question qualitatively. When looking at a fuel bed's response to incident heating, there are two major parameters that can be changed: the total amount of energy deposited to a fuel bed and the rate at which energy is deposited. Intuitively, increasing both of these parameters should increase the probability of ignition. Results from literature support this reasoning where increasing the energy deposition and/or the heat transfer rate increased the probability of ignition. Table~\ref{tab:binary} provides a summary of experimental efforts that have considered the ignition probability of fuel beds. For each study in Table~\ref{tab:binary} the fuel bed material, heat source, and parameters varied are outlined. The relative increase or decrease in ignition probability for each parameter change is also identified.

    %   \begin{table}[htbp]
    %          \rowcolors{2}{gray!25}{white}
    %          \caption{Summary of studies that have considered ignition probability of fuel beds. $\downarrow$ signifies an decrease in value and $\uparrow$ an increase}
    %             \centering
    %             \begin{tabular}{N p{1.5cm} p{3cm} p{3cm} p{4.5cm} p{0.7cm}}
    %             \rowcolor{gray!50}
    %                 \multicolumn{1}{|c}{ID} &  Fuel Bed & Heat Source & Parameter(s) & $\uparrow$ Ignition Probability & Ref  \\
    %                  \hline
    %                 \label{tabrow:wang2017}&             pine needles & metal  & FMC, $T_{ember}$, $U_{wind}$ \nomenclature{$U_{wind}$}{Air velocity over the fuel bed} & $\downarrow$ FMC, $\uparrow T_{ember}$ \nomenclature{$T_{ember}$}{Temperature of the ember}, $\uparrow U_{wind}$  & \cite{Wang2017}\\

    %                 \label{tabrow:urban2017}&            powdered grass & metal &   $T_{ember}$, $d_{ember}$ \nomenclature{$d_{ember}$}{Ember diameter} & $\downarrow d_{ember}$ requires $\uparrow T_{ember}$ & \cite{Urban2017}\\

    %                 \label{tabrow:Fernandez-Pello2015}&  cellulose & metal &     $T_{ember}$, $d_{ember}$ & $\downarrow d_{ember}$ requires $\uparrow T_{ember}$ & \cite{Fernandez-Pello2015}\\

    %                 \label{tabrow:ellis2015}&            forest litter & glowing or flaming wood & flaming or glowing ember, FMC,  $U_{wind}$ &  $\uparrow U_{wind}$, $\downarrow$ FMC & \cite{Ellis2015}\\

    %                 \label{tabrow:santoni2014}&          pine needles & radiative & $A_{s}/V$, permeability & $\uparrow A_{s}/V$ \nomenclature{$A_{s}$}{Surface area of the ember} \nomenclature{$V$}{Ember Volume}, $\uparrow$ permeability & \cite{Santoni2014}\\

    %                 \label{tabrow:reszka2012}&           nylon, PMMA & radiative & heating rate & $\uparrow$ heating rate & \cite{Reszka2012}\\

    %                 \label{tabrow:hadden2011}&           cellulose & metal & $T_{ember}$, $d_{ember}$ & $\downarrow d_{ember}$ requires $\uparrow T_{ember}$ & \cite{Hadden2011}\\

    %                 \label{tabrow:ellis2011}&            forest litter & flaming bark & FMC, $U_{wind}$, glowing mass &  $\downarrow$ FMC, $\uparrow$ glowing mass, $\uparrow U_{wind}$ & \cite{Ellis2011}\\

    %                 \label{tabrow:ganteaume2009}&        forest litter & flaming wood & $\rho$, FMC, species & $\downarrow$ FMC, $\downarrow \; \rho$ , species effect & \cite{Ganteaume2009}\\

    %                 \label{tabrow:plucinski2008}&        forest litter & cotton balls, aerial incendiary & FMC, species, $U_{wind}$ &  $\downarrow$ FMC, species effect, wind effect,  $\uparrow E_{ember}$\nomenclature{$E_{ember}$}{Energy content of ember}& \cite{Plucinski2008}\\

    %                 \label{tabrow:manzello2006}&         litter, paper, crevices & glowing/flaming wood & $U_{wind}$, FMC, $d_{ember}$, $N_{ember}$ \nomenclature{$N_{ember}$}{Number of embers} & material effect, flaming/glowing, $\downarrow$ FMC,  $\uparrow N_{ember}$ & \cite{Manzello2006}\\

    %                 \label{tabrow:manzello2006a}&        various mulches & glowing/flaming wood &  $U_{wind}$, FMC, $d_{ember}$ & material effect, flaming/glowing, $\downarrow$ FMC, $\uparrow U_{wind}$,  $\uparrow N_{ember}$ & \cite{Manzello2006a}\\

    %                 \label{tabrow:yang2003}&             wood plates & radiative & heat flux & $\uparrow$ heat flux & \cite{Yang2003}\\

    %                 \label{tabrow:delichatsios}&         wood plates & radiative & heat flux & $\uparrow$ heat flux & \cite{Delichatsios2003}\\

    %                 \label{tabrow:dimitrakopoulos2001a}&  various leaf species & radiative & FMC, species & species dependence, $\downarrow$ FMC & \cite{Dimitrakopoulos2001a}
    %             \end{tabular}
    %             \label{tab:binary}
    %             \end{table}

    %     Experiments conducted by Fernandez-Pello et al.~\cite{Fernandez-Pello2015}, in which hot metal particles were dropped onto cellulose, indicated the determining factor for ignition was proportional to the energy of the firebrand. More specifically, ignition was observed if the amount of thermal energy available from the particle delivered enough energy such that sufficient pyrolyzates were generated. This relationship manifests as a parabolic relationship between the particle diameter and energy of the particle at the ignition boundary as shown in Figure~\ref{fig:energy_diameter}. 
    %         \begin{figure}[hpbt]
    %             \centering
    %             \includegraphics[width=0.5\textwidth]{Figures/particle_ign_boundary.png}
    %             \caption{Observed ignition boundary, quantified by particle energy, as a function for cellulose powder for various materials and particle sizes from Fernandez-Pello et al.~\cite{Fernandez-Pello2015}.}
    %             \label{fig:energy_diameter}
    %         \end{figure}
    %     Similar results for hot particles were obtained for pine needles by Wang et al.~\cite{Wang2017}, for powdered grass by Urban et al.~\cite{Urban2017}, and again with cellulose by Hadden et al.~\cite{Hadden2011}. While similar trends have been observed across a range of materials using similar energy delivery methods, efforts to create a model to predict ignition boundaries have not been successful. The difficulties preventing an ignition model have stemmed from the lack of knowledge about the ember fuel bed interface, thermal properties of the fuel bed, and difficulties in determining particle heat losses to ambient.

    %     While observations informing trends that increased energy and heat transfer increase  probability are important, the next step is to quantify the transition or critical values for parameters observed to control the trends. A universally predictable critical value has not yet been determined but, predictive capabilities for single conditions or experimental apparatus have been determined. More detail on what has been learned about the impact of energy deposition and effects of energy deposition on pyrolysis is contained in the following sections.

    %     \section{Fuel Moisture Content}
    %     The presence of water in fuel bed particles, quantified as fuel moisture content, impacts the ignition process in multiple ways. Since the vaporization of water occurs at a lower temperature than pyrolysis, the water in a fuel particle must be evaporated before pyrolysis can occur. This creates a large energy sink for the energy imparted to the fuel bed from an ember or firebrand, thus increasing the amount of energy needed to ignite a fuel bed~\cite{Hurley2016}. Studies have universally shown this to be the case as is described in Table~\ref{tab:binary} rows~\ref{tabrow:wang2017}, \ref{tabrow:ellis2015}, \ref{tabrow:ellis2011}, \ref{tabrow:ganteaume2009}, \ref{tabrow:plucinski2008}, \ref{tabrow:manzello2006}, \ref{tabrow:manzello2006a}, and~\ref{tabrow:dimitrakopoulos2001}. This trend indicates that fuel moisture content is a parameter to be added as a lumped constant to the minimum energy needed to ignite a fuel bed. However, water in the fuel has additional impacts on the pyrolysis and ignition processes. Computational efforts have shown that increasing the moisture content of the fuel decreases both the peak mass loss rate and peak heat release rate of a fuel sample\cite{Shotorban2018}. The decrease in the heat release rate and mass loss rate may have significant effects on the chemical composition of the pyrolysis gases, as was observed by Furgeson et al.~\cite{Ferguson2013} where an increased moisture content shifted the \ce{O2} and \ce{H} concentrations in flames burning pyrolysis gases. It should be noted that this effect was observed for the piloted combustion of pyrolysis gases, but increasing amounts of \ce{H2O} would likely have a similar impact on the auto-ignition of pyrolysis gases. The implications of these effects are discussed further in Section~\ref{subsec:pyrolysis}, as they relate more directly to pyrolysis of the fuel and gas dynamics above the fuel bed.

    %     In addition to acting as an energy sink for evaporation, increasing the fuel moisture content alters the thermal conductivity of a fuel bed. Tests for the thermal conductivity of wheat showed that a 28\% increase in wheat moisture content increased the thermal conductivity by ~40\%\cite{Tavman1998}. The increased heat diffusion rates would lead to lower temperatures and temperature gradients in the fuel bed, changing the mass flux rate of pyrolyzates departing the fuel bed and the composition of said pyrolyzates. The relative magnitude of increased moisture content in other fuel bed materials is unclear, but a ~40\% shift in thermal conductivity would have significant impacts on the heat diffusion rates from a firebrand delivering energy to the fuel bed.
    %     Perhaps the most important knowledge gained from the extensive study of fuel moisture content and its impacts is that due to the numerous scales across which ignition of fuel beds occur it is essential to evaluate the impacts of each parameter across all scales since the phenomena may be more closely interdependent than initially observed.

    %     \section{Species and Morphology}
    %     One of the factors that sets the ignition of fuel beds apart from ignition of other materials is the vast array of species and particle morphology encountered in nature. For example, in a forest with a relatively uniform litter layer, the particle size distribution may range from small pieces of grass to twigs that are an order of magnitude larger. Even a layer of pine needles on a roof will have a distribution of sizes, shape, orientation, and bulk density. On the scale of the fuel bed, the differences in shapes and sizes of the particles ensure that every fuel bed is unique, making repeatability while studying fuel bed ignition difficult at best.

    %     Implications for fuel bed uniqueness is twofold. First, the differences in particle shape, e.g., pine needles vs oak leaves, cause the heat transfer properties of fuel beds to be highly variable. Second, each species has different composition and fuel beds may consist of many species. Having fuel bed particles of different compositions and morphology in a fuel bed obfuscates the relative effects of individual particle thermal conductivity and inter-particle heat transfer due to contact area and particle shape. Furthermore, a fuel bed with multiple species will undergo a wide array of chemical processes in a single fuel bed when pyrolyzates are generated. Samples of the same fuel have also been shown to have different pyrolyzate composition under differing heating conditions~\cite{Gauthier2013}. For example, decreasing the heating source from 1050$^{\circ}$C to 450$^{\circ}$C reduced the elemental composition of hydrogen in the pyrolysis products by 23\% and increased the elemental composition of oxygen by 28\% on a \% mole basis. The inherent variability in natural fuel bed has limited observations to qualitative trends.

>>>>

\chapter{Background}
\label{chap:literature}
    This chapter provides a discussion of the background and motivation of this work in further depth than the manuscripts presented later in this dissertation. Here, a general overview of ignitions as a threat to structures and fire spread and the current state of knowledge and predictive capabilities is communicated. Each manuscript includes a relevant and more detailed literature review pertaining to the specific objective of the chapter.

@@

\section{The Problem of Ignition}
    Ignition of a home in a WUI fire is attributed to three primary mechanisms; direct flame contact, ignition by radiative heating, and ignition by firebrands. Of these methods, ignition by firebrands is considered to be a primary threat to homes~\cite{Suzuki2021, Mell2010, Manzello2018FORUMResearch}. Case studies of WUI fires have consistently attributed structural losses to firebrand ignition~\cite{Westhaver2017WhyDisaster, Roberts2021} with one study attributing at least 2/3 of home losses to firebrands~\cite{Mell2011}. With the majority of ignitions occurring due to firebrand ignition, it is imperative to understand the mechanism of firebrand ignition to reduce home losses. An analysis of structures burned in California wildfires from 2013 to 2018 found that many of the homes lost would have been considered "fire-safe" by current standards but were still ignited by firebrands~\cite{Syphard2019}. Codes, standards, and best practices built upon a better understanding of ignition by firebrands would help provide better resilience to firebrand attacks~\cite{Manzello2020} reducing the number of homes lost in wildfires. Examples of these changes may include adjustments in building and landscaping materials favoring the selection of harder to ignite species or housing geometry constraints to reduce the collection of flammable materials or firebrands. Without a comprehensive knowledge of the ignition process, creating the codes and standards to reduce the risk of home loss is challenging at best and may even be counterproductive.

    The general mechanism for firebrand ignition is considered to be a three-step process that starts within the fire itself, often at a significant distance from the ignition point. First, a firebrand is generated as burning material breaks off or is lofted into the air. Second, the firebrand is then lofted through the air and transported to the ignition site. Third, a firebrand lands on a recipient fuel bed, and then ignition occurs if the criteria for ignition are met~\cite{Babrauskas2003}. When considering the possibility of ignition during a fire event, all three processes must be considered as each process significantly influences the properties of a firebrand landing on a fuel bed. For example, the number, energy content, and size of firebrands that may land on a fuel bed are significantly influenced by the surrounding foliage~\cite{Hudson2020EffectsScale} and moisture content of the foliage~\cite{Adusumilli2021FirebrandShrub} which introduces significant variability in ignition sources a fuel bed may encounter. Similarly, the winds transporting a firebrand significantly influence the path of travel and time of flight, which influences the energy content~\cite{Sardoy2007, MatvienkoOVA2016} and localized winds around structures may lead to firebrand accumulation~\cite{Suzuki2020a, Suzuki2017a, Suzuki2015}. Unfortunately, the generation and ignition processes are significantly less understood than the transport processes~\cite{Manzello2020}. The work contained in this dissertation is focused on the ignition process of the fuel bed by a firebrand and does not consider the generation of firebrands. For further information on firebrand generation processes, the reader is referred to review papers considering this process by Fernandez-Pello~\cite{Fernandez-Pello2017}, Manzello et al.~\cite{Manzello2020} and Babrauskas~\cite{Babrauskas2003}.

    Firebrands have been identified as a source of structure loss as early as the 1600s~\cite{Suzuki2020} however, studies of firebrand ignition have only begun recently. Perhaps one of the first studies published to specifically address firebrands within the scope of structure loss was conducted by Waterman in 1969~\cite{Waterman1969ExperiemntalGeneration}. However, research remained sparse until the early 2000s. Early research and a significant portion of current research have been conducted by dropping combusting firebrands~\cite{Ellis2011, Ellis2015, Ganteaume2009, Manzello2006, Manzello2006a} (either flaming or glowing) or hot metal particles~\cite{Wang2017, Urban2017, Fernandez-Pello2015, Hadden2011} onto fuel beds in varying configurations that simulate conditions in a wildfire and observing ignition or no ignition outcomes. These works have identified some parameters that appear to control ignition but models created from these studies only provide qualitative predictions of ignition. The lack of models of quantitative accuracy was suggested by Finney et al. in 2013, who identified the lack of a fundamental understanding of what processes occur and how they interact as a primary gap in the knowledge of ignition~\cite{Finney2013}. A similar need was identified by Manzello et al. in 2018~\cite{Manzello2018} and again in 2020~\cite{Manzello2020}. As fires increase in severity and the WUI continues to expand, the number of homes lost to wildfires is likely to increase in the near future significantly. Thus, the need for accurate predictions of ignition has never been higher.

\section{Modeling Efforts}
    Creating a model with quantitative accuracy of ignition requires a thorough understanding of the processes and relative importance of each process leading up to ignition. Figure~\ref{fig:ignitionDiagram} illustrates the different factors that can influence the ignition process. To better understand how these parameters affect the ignition process can be considered as a series of three processes. The first is the heat transfer to the fuel bed from the firebrand. The second is the pyrolysis of the fuel bed material and the subsequent release of pyrolyzates into the air above the fuel bed. The third step is the mixing of the pyrolyzates with air above the fuel bed and subsequent gas-phase ignition. It is important to note that all three of these processes may occur simultaneously throughout the ignition process and are often interconnected. For example, as heat is transferred from the firebrand to the fuel bed, the thermal conductivity of the fuel bed changes with temperature both before and after pyrolysis begins~\cite{Fjellerup2003}. Furthermore, the heat transfer mode and thermal resistance between the firebrand and fuel continuously change as mass is released due to pyrolysis, which may influence ignition~\cite{Yang2016EffectParticle}. With the interconnection of the processes that occur during ignition and the inherent variability of fuel beds and biomass fuels in general, the quantification of parameters that influence ignition is quite challenging. Despite these challenges, results of previous research have identified influential parameters that have consistently been observed to control ignition. This work considers these parameters as belonging to one of three categories, namely, thermo-physical properties of the fuel bed, energy supplied by the firebrand, and environmental conditions near the ignition point. Detailed reviews of each category are provided in the relevant manuscripts and are not reproduced here. The remaining sections of this review address the predictive models of ignition that have been created from the current state of knowledge. The accuracy and applicability of each model are discussed in turn, and the remaining challenges are identified as they have motivated the objectives of this work. 
        \begin{figure}[hpbt]
            \centering
                \includegraphics[width=\textwidth]{Figures/dissertationBed.pdf}
            \caption{Illustration of key features of fuel bed ignition and parameters/characteristics that may influence ignition. The black region near the firebrand represents char formed as the result of pyrolysis.}
            \label{fig:ignitionDiagram}
        \end{figure}

@@

        \begin{equation}
            r_{cr} = \delta_{cr}\sqrt{\frac{k}{\rho A \Delta H}\frac{RT^{2}_{p0}}{E} exp\left(\frac{E}{RT_{p0}}  \right)}
            \label{eqn:hotspot}
        \end{equation}
    In this equation, $\delta_{cr}$ is defined as the critical Frank-Kamenetskii hot spot parameter and is determined from calculations~\cite{Goldshleger1973IgnitionDimensions}. $\delta_{cr}$ is dependent on the ratio of volumetric heat capacities of the ignition source and the material, the activation energy of the material, and the temperature of the hot particle. In Equation~\ref{eqn:hotspot} $k, \rho A, E$, and $\Delta H$ are the thermal conductivity, density, pre-exponential factor, activation energy, and heat of combustion of the fuel bed material, respectively. $R$ is the universal gas constant, and $T_{p0}$ is the initial temperature of the heat source. Ultimately, Equation~\ref{eqn:hotspot} defines the particle energy needed to raise the temperature of a fuel bed region such that thermal runaway and subsequent ignition occurs. The hot spot theory has some limitations that prevent quantitative predictions of ignition in a wildfire setting. A discussion of the shortcomings of the hot spot theory is presented by Thomas~\cite{Thomas1965} and more recently presented by Hadden et al.~\cite{Hadden2011} with respect to wildfires. The two most significant shortcomings addressed by the previous works are neglecting heat losses to the surroundings and considering the pyrolysis and ignition phenomena as single-step, infinitely fast reactions. Hadden et al. attribute these shortcomings as a significant source of error for quantitative ignition predictions. This assertion is reinforced in a review by Fernandez-Pello~\cite{Fernandez-Pello2017} who concluded that in instances where hot spot theory has been applied, only qualitatively accurate predictions of ignition had been obtained. A similar model to hot spot theory but with a more comprehensive inclusion of heat loss from the firebrand was developed by Yin et al.~\cite{Yin2014}. A correlation between fuel properties, the energy transferred to the fuel bed, and the time to ignition was proposed and is shown in Equation~\ref{eqn:yinMC}.
        \begin{equation}
            \sqrt{t_{ig}} \approx \frac{q\sqrt{\frac{\rho_{f}k}{c_{p}}}}{\frac{\rho Z \Delta H_c}{t_b}- h_T(T_f - T_0)}
            \label{eqn:yinMC}
        \end{equation}
    Comparing the correlations presented in Equations~\ref{eqn:hotspot} and \ref{eqn:yinMC}, Equation~\ref{eqn:yinMC} includes properties of the fuel bed ($\rho$, $k$, $c_p$, $Z$, $T_0$) and  firebrand ($\rho_f$, $\Delta H_c/t_b$, $T_f$) as well as the energy required for ignition ($q$) mirroring the Frank-Kamenetskii critical hot spot diameter. The addition of heat losses to the ambient is the most significant difference from the hotspot parameter and addresses a key shortcoming of the hot spot theory. However, the model does not consider the combustion of the fuel bed other than the critical energy for ignition, leaving the second key gap for the hot spot theory unaddressed. Perhaps due to the omission of combustion, this model suffers a similar fate as the hot spot theory and follows qualitative trends but is not quantitative. Ultimately, the hot spot theory and similar models do not capture the complexity and variability inherent to ignition~\cite{Manzello2020}.

@@

    Proposed ignition models that provide greater accuracy than the hot spot and critical mass flux models do so at the expense of complexity and computational cost. The majority of these models and modeling efforts use a two-part approach to join a pyrolysis-specific solver to address the chemical kinetics of an existing or custom computation fluid dynamics solver. This approach requires selecting a combination of three parts, each of which has limitations and considerations which may affect the accuracy of ignition predictions. First, an appropriate chemical kinetic mechanism must be chosen based on the computational resources available. For example, many early works~\cite{DiBlasi1996} and some more recently~\cite{Ding2015, Torero2020} have considered pyrolysis to occur as a single step process or as a process of few reactions (approximately ten or fewer). Newer kinetic mechanisms have been developed to more completely capture the processes of pyrolysis and are much more complex. Perhaps the most comprehensive mechanisms are extended forms of the mechanism developed by Ranzi et al.~\cite{Ranzi2008}. Additions to this mechanism have expanded the method to include extractives to cover a greater variety of biomass~\cite{Debiagi2015} and the addition of reactions to address gas-phase reactions previously not considered~\cite{Dhahak2019}. For a more detailed discussion of the progression and challenges of biomass pyrolysis, the reader is referred to review articles by Di Blasi~\cite{DIBLASI199371} and more recently by Haberle et al.~\cite{Haberle2017}.

   One of the most recent and most promising tools for ignition modeling and pyrolysis is Gpyro which is a pyrolysis solver specifically designed to handle both the condensed (solid) and gas phase of biomass combustion~\cite{Lautenberger2009a}. Gpyro is often coupled with the Fire Dynamics Simulator (FDS)~\cite{FDSManual}, an LES solver specifically designed to capture physics relevant to fires, to form a complete computational package. The combination of Gpyro and FDS has been utilized to observe combustion sensitivities and ignition sensitivities. For example, a study of the effects of moisture content on ignition using Gpyro/FDS suggested that moisture in the fuel affected ignition, gas-phase combustion, and pyrolysis of the fuel~\cite{Yashwanth2016}. Other studies using Gpyro/FDS have been conducted to predict ignition and observe controlling parameters~\cite{Shotorban2018, Fernandez-Pello2015, Lautenberger2009}. Unfortunately, the Gpyro/FDS modeling requires significant computational expenditure. The high computational cost of tools like Gpyro/FDS prevents them from practical application across the vast number of configurations and possible ignition threats present in the wildland urban interface.

    To balance the need for accuracy and computational cost, a usable and accurate model is anticipated to be between the critical mass flux and the full CFD model with respect to complexity. The understanding of the effects (e.g., environmental conditions, fuel bed properties.) that are most influential to ignition is not complete enough to develop such a model. Bridging this knowledge gap by identifying and quantifying which parameters are the most influential to ignition or non-ignition outcome is the overall goal of this work. The identification of the most influential parameters is anticipated to enable the creation of a model that can predict ignition across a variety of fuel beds, firebrands, and environmental configurations.
